---
title: "Extra Sum-of-Squares"
institute: "**Jam TBD**"
---

```{r}
#| include: false
set.seed(2112)
```


::: {.content-visible unless-profile="book"}
## Preamble

### Announcements

- Check discussions!
:::

## Introduction

### Today's Main Idea

If you add or remove predictors, the variance of the residuals changes!

\pspace

- As always, we ask if it's a "big" change.\lspace
- Different predictors have a different effect on the residuals.

\pspace\pause

Which predictors have a meaningful (significant?) effect on the residuals?

### Sum-of-Squares due to Regression

- Since SSE = SST - SSReg and SST never changes, we're focusing on SSReg.\lspace
- Recall: SSReg is the variance of the line itself!

$$
SSReg = \sum_{i=1}^n(\hat y_i - \bar y)
$$

### SSReg in two different penguin models

In the `penguins` data, we're determining which predictors are associated with body mass.

- $SS_1$ = SSReg for model 1
    - $\texttt{body\_mass\_g} = \beta_0 + \beta_1 \texttt{flipper\_length\_mm} + \beta_2 \texttt{bill\_length\_mm} + \beta_3 \texttt{bill\_depth\_mm}$
- $SS_2$ = SSReg for model 2
    - $\texttt{body\_mass\_g} = \beta_0 + \beta_1 \texttt{flipper\_length\_mm} + \beta_2 \texttt{bill\_length\_mm}$

Note: M2 is **nested** in M1 - M1 has all the same predictors and then some.

\pspace

::: {.callout-important}
$\beta_1$ in the first model is **different** from $\beta_1$ in the second model.
:::

### Extra Sum-of-Squares

If M2 is **nested** within M1, \emph{e.g.}:

- M1: $\texttt{bodymass} = \beta_0 + \beta_1 \texttt{flipperlength} + \beta_2 \texttt{billlength} + \beta_3 \texttt{billdepth}$
- M2: $\texttt{bodymass} = \beta_0 + \beta_1 \texttt{flipperlength} + \beta_2 \texttt{billlength}$

Then the Extra sum of squares is defined as:
$$
SS(\hat\beta_3 | \hat\beta_0, \hat\beta_1, \hat\beta_2) = S_1 - S_2
$$

\pspace

Convince yourself that S1 > S2.

### Special Case: Corrected Sum-of-Squares

We've already seen this notation:
$$
SSReg(\hat\beta_0) = n\bar{\underline y}^2
$$
and
$$
SSReg(corrected) = \hat{\underline\beta}^TX^T\underline y - n\bar{\underline y}^2 = S_1 - S_2
$$
where $S_2$ is the sum-of-squares for the null model!

### Unspecial Case: Correction doesn't matter!

Consider $S_{1c}$ and $S_{2c}$, the corrected versions of $S_1$ and $S_2$. Then

\begin{align*}
S_{1c} - S_{2c} = (S_2 - n\bar{\underline y}^2) - (S_1 - n\bar{\underline y}^2) = S_1 - S_2
\end{align*}

\pspace

In other words, the correction term doesn't matter. 

This is useful because R outputs the corrected versions.

### Unspecial Case: SSReg versus SSE doesn't matter!

Consider $SSE_1$ and $SSE_2$. Since SST is the same for both models,

\begin{align*}
SSE_2 - SSE_1 = (SST - S_1) - (SST - S_2) = S_2 - S_1
\end{align*}

Notice that the order is switched, which is fine.

### ANOVA Tests for ESS

Consider the models:

- M1: $\texttt{bodymass} = \beta_0 + \beta_1 \texttt{flipperlength} + \beta_2 \texttt{billlength} + \beta_3 \texttt{billdepth}$
    - $df_1 = 4$
- M2: $\texttt{bodymass} = \beta_0 + \beta_1 \texttt{flipperlength} + \beta_2 \texttt{billlength}$
    - $df_2 = 3$

\pspace

If we choose $H_0: \beta_3 = 0$ in model 1, then
$$
\frac{S_1 - S_2}{(4 - 3)s^2} \sim F_{1,\nu}
$$

where $s^2$ is the error variance (MSE) in the larger model with degress of freedom $\nu = df_1$.

This is almost identical to the F-test for only one predictor (with one important difference).

### In General

If M1 has $p$ df, M2 has $q$ df, and one is nested in the other, then $\nu = \max(p, q)$ and

$$
\frac{S_1 - S_2}{(p - q)s^2} \sim F_{|p-q|,\nu}
$$

\pspace

Note that it doesn't matter which is nested: $S_1 - S_2$ has the same sign as $p-q$, so it's always positive.

### Omnibus Tests for Multiple Predictors

Suppose we want to test if *any* bill measurement is useful. 

- Bill length and depth are highly correlated - marginal CIs won't be valid.
- Confidence Regions are hard (and only work in 2D)

\pspace

Instead, we can use the ESS to test for a subset of predictors!

- M1: $\texttt{bodymass} = \beta_0 + \beta_1 \texttt{flipperlength} + \beta_2 \texttt{billlength} + \beta_3 \texttt{billdepth}$
- M2: $\texttt{bodymass} = \beta_0 + \beta_1 \texttt{flipperlength}$

$S_1 = S_2$ is equivalent to $\beta_2 = \beta_3 = 0$, and it accounts for their covariance!

\pspace

If significant, then at least one of $(\beta_2, \beta_3)$ is not 0.

### In R

```{r}
#| eval: true
#| echo: true
library(palmerpenguins)
peng <- penguins[complete.cases(penguins),]
m1 <- lm(body_mass_g ~ flipper_length_mm + bill_length_mm + bill_depth_mm,
    data = peng)
m2 <- lm(body_mass_g ~ flipper_length_mm,
    data = peng)
anova(m1, m2) |> knitr::kable()
```

### Next time

- When to check ESS\lspace
- How to check all ESS\lspace
- What is R's `anova()` function even doing??


