---
title: "The Hat Matrix"
institute: "Jam: **Residual** by Trash."
---

```{r}
#| include: false
set.seed(2112)
```


::: {.content-visible unless-profile="book"}
## Preamble

### Announcements



:::

## Le Chapeau

### The Hat Matrix

$$
H = X(X^TX)^{-1}X^T
$$

The hat matrix projects $Y$ onto $\hat Y$, based on $X$.

- $\hat Y = HY$
    - $\hat Y_i = h_{ii} Y_i + \sum_{j\ne i}h_{ij}Y_j$

\pspace

In other words, $h_{ii}$ determines the *influence* of the observed point $y_i$ on it's own prediction.

### Variance-Covariance Matrix of $\hat{\underline\epsilon}$

Just like $\beta_0$ and $\beta_1$, each sample results in different $\underline{\hat\epsilon}$.

Across samples, we have:
$$
\underline{\hat\epsilon} - E(\underline{\hat\epsilon}) = (I-H)(Y-X\underline{\beta}) = (I-H)\underline{\epsilon}
$$
and therefore:
\begin{align*}
V(\underline{\hat\epsilon}) &= E([\underline{\hat\epsilon} - E(\underline{\hat\epsilon})][\underline{\hat\epsilon} - E(\underline{\hat\epsilon})]^T)\\
&= [I-H]E(\underline{\epsilon}\underline{\epsilon}^T)[I-H]^T\\
&= [I-H]\sigma^2[I-H]\\
&= [I-H]\sigma^2
\end{align*}
where we used the idempotency and symmetry of $I-H$.


### The variance of a residual

Given that $V(\underline{\hat\epsilon}) = (I-H)\sigma^2$,
$$
V(\hat\epsilon_i) = (1-h_{ii})\sigma^2
$$

- The variance of the residual depends on how much $Y_i$ influences it's own estimate.
    - High influence = low variance.

\pspace

The correlation between residuals is:
$$
\rho_{ij} = \frac{Cov(\hat\epsilon_i, \hat\epsilon_j)}{\sqrt{V(\hat\epsilon_i)V(\hat\epsilon_j)}} = \frac{-h_{ij}}{\sqrt{(1-h_{ii})(1-h_{jj})}}
$$

- The covariance is negative! A large residual tells us there are small residuals.
    - "Large" and "small" are relative

### More H Facts

1. $SS(\hat{\underline\beta}) = \hat{\underline\beta}^TX^TY = \hat Y^TY = Y^TH^TY = Y^TH^THY = \hat Y^T\hat Y$
    - We used the facts $H^T= H^TH$ and $\hat Y = HY$.\lspace
2. $\sum_{i=1}^nV(\hat Y_i) = trace(H\sigma^2) = p\sigma^2$
    - $p$ is the number of parameters.
    - Proof is homework\lspace
3. $H1 = 1$ if the model contains a $\beta_0$ term.
    - $1$ is a *column* of 1s, not identity matrix.
    - Proof on next slide.
    - $h_{ii} = 1 - \sum_{j\ne i}h_{ij}$
        - Note that $h_{ij}\in[-1, 1]$.

::: {.content-visible when-profile="book"}
The first one is a little hard to believe! How can $\hat Y^TY = \hat Y^T\hat Y$ when $\hat Y \ne Y$???

```{r}
mylm <- lm(mpg ~ wt, data = mtcars)
X <- model.matrix(mylm)
Y <- mtcars$mpg
Yhat <- predict(mylm)
H <- X %*% solve(t(X) %*% X) %*% t(X)
beta <- as.numeric(coef(mylm)
t(beta) %*% t(X) %*% Y
t(Yhat) %*% Y
t(Y) %*% H %*% Y
t(Y) %*% t(H) %*% H %*% Y
t(Yhat) %*% Yhat
```



:::

### Proof that $H1 = 1$

Note that $HX = X$ (as proven on A1).

1. The first column of $X$ is all ones ($\beta_0$ term).
    - $[X]_{i1} = 1$\lspace
2. Therefore $[HX]_{i1}$ is a column of ones.
    - Every row of $H$ times the column of 1s in $X$ results in a column of ones.\lspace
3. $[HX]_{i1}$ is every row of $H$ times the first column of $X$.

\pspace

The first column of $X$ is 1s, which is equal to the first column of $HX$, which is $H$ times a column of ones.

In other words, $H1 = 1$

## Studentized Residuals

### Internally Studentized (not ideal)

How do you measure the size a residual?

\pause

Divide by the variance, of course!

\space

We know that $V(\hat \epsilon_i) = (1-h_{ii})\sigma^2$, and
$$
s^2 = \frac{\sum_{y=1}^n(y_i-\hat y_i)^2}{n-p} = \frac{SSE}{df_E} = MSE
$$
is an estimate of $\sigma^2$. Then,
$$
r_i = \frac{\hat\epsilon_i}{\sqrt{s^2(1-h_{ii})}}
$$
is called the **internally studentized residual**.

### "Internally" "Studentized"

Note that
$$
s^2 = \frac{\sum_{i=1}^n(y_i-\hat y_i)^2}{n-p} = \frac{\sum_{i=1}^n(\hat\epsilon_i)^2}{n-p}
$$
and therefore
$$
r_i = \frac{\hat\epsilon_i}{\sqrt{s^2(1-h_{ii})}} = \frac{\hat\epsilon_i}{\sqrt{(\hat\epsilon_i^2 + \sum_{j\ne i}\hat\epsilon_j^2)(1-h_{ii})/(n-p)}}
$$

- If $\hat\epsilon_i$ is large, then $s^2$ is large.
    - If $s^2$ is large, then $r_i$ is small!
    - "**Internally**": the variance includes the residual of interest.\lspace
- "**Studentized**" because Student made it popular.\lspace
- Often called "standardized".

### Externally Studentized Step 1

Like adding/removing predictors and checking the change in SS, we can add/remove points!

\pspace

1. Calculate SS
2. Remove the first point. Estimate the model again and calculate new SS.
3. Add the first point back, remove the second. Estimate the model again and check the SS.
4. ...

\pspace

For each point, we have an estimate of the variance **without** itself.

### Externally Studentized Step 2

Skipping the math,
$$
s^2_{(i)} = \frac{(n-p)s^2 - \hat\epsilon_i^2/(1-h_{ii})}{n-p-1}
$$
is the variance of the residuals **without** observation $i$.

\pspace

- The influence tells us how much a point influenced the model
    - We can see what happened without it
    - No need to re-estimate the model!!!

### Externally Studentized Residuals

Use $s^2_{(i)}$ in place of $s^2$.
$$
t_i = \frac{\hat\epsilon_i}{\sqrt{s_{(i)}^2(1-h_{ii})}} \sim t_{n-p-1}
$$

- Follows a $t$ distribution!
    - Larger than 2 is suspect, 3 is definitely an outlier!
- A large $t_i$ is large *relative to the other residuals*
- Usually just called "Studentized"\lspace

Most software uses Studentized residuals for plots/diagnostics!



## Cook's Distance

### Better Measures of Influence

The hat matrix is intepreted as influence, but it has problems.

- $y_i$'s influence on *it's own* prediction,
    - *given all other points.*
- $0 \le h_{ii} \le 1$
    - What is a "big" influence?
- How do you explain $h_{ii}$ to nonstatisticians?

\pspace

A better measure is how much the predicted value changes with/without the obs.

### Cooks Distance: Change in $\hat y_i$.

$$
D_i = \frac{\sum_{i=1}^n(\hat y_i - \hat y_{(i)})^2}{ps^2}
$$

- $\hat y_i$ is the predicted value of $y_i$ when all data are considered.
- $\hat y_{(i)}$ is the predicted value of $y_i$ when observation $i$ is removed.
- $s^2$ is the MSE of the model with all of the data.
- $p$ is the number of parameters
    - $D_i$ decreases as $p$ increases!

\pspace

Again, this *would* involve re-fitting the model $n$ time (one for each obs).

### Cook's Distance: Alternate Form

$$
D_i = \left[\frac{\hat \epsilon_i}{\sqrt{s^2(1-h_{ii})}}\right]^2\frac{1}{p}\left[\frac{h_{ii}}{1 - h_{ii}}\right] = r_i^2\frac{1}{p}\frac{\text{variance of $i$th predicted value}}{\text{variance of $i$th residual}}
$$

\pspace

- Again, we can use $H$ rather than re-fitting the model.\lspace
- Cook's distance is a modification of the *internally* studentized residual.
    - Variances are based on same "deletion" idea as studentized.\lspace
- Ratio of Variances!
    - $F$ distribution, mean approaches 1 for large values of $n$
        - Cooks Distance of larger than 1 is suspect.

### Next Class

\centering
Plots!

::: {.content-visible unless-profile="book"}

## Participation

### Q1

Which does *not* describe a residual?

\pspace

a. The observed difference between the measured value $y_i$ and the one we predict with $\hat y_i = X\hat{\underline\beta}$.
b. Given the true relationship $Y = X\underline\beta$, residuals are deviations that cannot be observed.
c. Errors that should be fixed.

<!--- C --->

### Q2

Which of the following is *not* true about the hat matrix ($H = X(X^TX)^{-1}X^T$)?

\pspace

a. $(I-H)(I-H)^T = (I-H)$
b. $\hat{\underline\beta} = (X^TX)^{-1}X^TY = X^{-1}HY$
c. $h_{ii} = 1 - \sum_{j\ne i}h_{ij}$
d. $H(I-H) = 0$

<!--- B --->

### Q3

Suppose that $\epsilon_i \stackrel{iid}{\sim} N(0,\sigma^2)$.

Which of the following statements is *false*?

\pspace

a. $V(\hat\epsilon_i) = (1 - h_{ii})\sigma^2$
b. $V(\epsilon_i) = \sigma^2$
c. $V(Y_i) = \sigma^2$
d. $V(\hat Y_i) \ge V(\hat \epsilon_i)$

<!--- D --->

### Q4

The difference between internally and externally studentized residuals is:

\pspace

a. "Internal" uses all of the observations, "external" uses all of the observations except $i$.
b. "External" uses all of the observations, "internal" uses all of the observations except $i$.

<!--- A --->

### Q5

Internally studentized residuals follow a $t$ distribution.

\pspace

a. True - they are a normal r.v. divided by a chi-square r.v.
b. False - they are a normal r.v. divided by a chi-square r.v., but the two are *not* independent.
c. False - They include a normal and a chi-square, but the $h_{ii}$ make this not follow distributional assumptions.
d. True - internally studentized residuals are a small modification to externally studentized residuals, which follow a $t$ distribution.

<!--- B --->

### Q6

Which of the following is *not* useful for detecting outliers?

\pspace

a. Standardized residuals
b. Studentized residuals
c. Cook's distance
d. $h_{ii}$

<!--- D --->

:::

::: {.content-visible when-profile="book"}

## Exercises

Suggested Ch08 Exercises: **A**, **C**.

1. In the proof that $V(\underline{\hat\epsilon_i}) = (1-h_{ii})\sigma^2$, we skipped a few steps. Fill in those steps.
2. Explain why "a large residual tells us there are small residuals". It is useful to know that $\sum_{i=1}^n\hat\epsilon_i = 0$.
3. Explain where each part of the formula for $\rho_{ij}$.
4. Prove that $\sum_{i=1}^nV(\hat Y_i) = trace(H\sigma^2) = p\sigma^2$. You may use the fact that the "trace" function allows you to move the last term to the front (i.e. you can "cycle" the entries): $trace(ABCD) = trace(DABC) = trace(CDAB) = trace(BCDA) \ne trace(BACD)$.
5. For the following data, calculate $h_{11}$, $\hat\epsilon_1$, $s_1$, $t_1$, and $D_1$ using R. Plot the data; comment on whether the first value is an outlier, and whether you would know this from each of the calculated values. 

```{r}
set.seed(2112)
x <- c(10, runif(49, 0, 10))
y <- c(50, 2 + 3 * x[-1] + rnorm(49, 0, 3))
mylm <- lm(y ~ x)
X <- model.matrix(mylm)
H <- X %*% solve(t(X) %*% X) %*% t(X)
h11 <- H[1,1]
```

6. For the data in the previous question, note that the mean of $x$ should be around 5. Move the first point to this x-value, keeping the y-value the same. First, make a guess at how do each of $h_{11}$, $\hat\epsilon_1$, $s_1$, $t_1$, and $D_1$ change, writing down your guess. Then calculate the values and see if you were right!


:::


