---
title: "The Hat Matrix 2"
institute: "**Jam TBD**"
---

```{r}
#| include: false
set.seed(2112)
```

## Participation Questions

### Q1

Which of these does *not* give the diagonal of the hat matrix?

1. `hatvalues(mylm)`
2. `diag(X %*% solve(t(X) %*% X) %*% t(X))`
3. `augment(mylm)$hat`

### Q2

Removing an outlier will *not* change:

1. The diagonal of the hat matrix.
2. The off-diagonal of the hat matrix.
3. SST
4. All of the above will change if we remove an outlier.

### Q3

A large residual means large influence.

1. True
2. False

### Q4

All entries in the hat matrix are between -1 and 1.

1. True
2. False

### Q5

Which plot corresponds to the model with the *most* predictors?

(All models are nested.)

```{r}
#| echo: false
#| eval: true
#| fig-height: 4.5
#| fig-width: 12
library(patchwork)
library(ggplot2)
library(mlbench)
library(broom)
data(Ozone)
theme_set(theme_bw())

Ozone <- Ozone[complete.cases(Ozone), ]

lm1 <- augment(lm(V4 ~ V6, data = Ozone))
lm2 <- augment(lm(V4 ~ V6 + V7, data = Ozone))
lm3 <- augment(lm(V4 ~ V6 + V7 + V5, data = Ozone))
lm4 <- augment(lm(V4 ~ V6 + V7 + V5 + V8 + V9 + V10, data = Ozone))

g1 <- ggplot(lm1) + aes(x = .fitted, y = .resid)
g2 <- ggplot(lm2) + aes(x = .fitted, y = .resid)
g3 <- ggplot(lm3) + aes(x = .fitted, y = .resid)
g4 <- ggplot(lm4) + aes(x = .fitted, y = .resid)

(g1 + g4) / (g3 + g2)+
    plot_annotation(tag_levels = '1') & geom_point(size = 2) &
    geom_smooth(method = "loess", formula = y ~ x, se = FALSE) &
    coord_cartesian(ylim = c(-20, 30)) 

```

### Q6

In the output of `augment()` in the `broom` package, the `.sigma` column refers to:

1. The MSE of the model.
2. The MSE of the model *if it were fit without the observation in that row*.
3. The variance of that residual, as calculated by $V(\hat\epsilon_i) = (1-h_{ii})s^2$.
4. The variance of that residual, as calculated by $V(\hat\epsilon_i) = (1-h_{ii})s_{(i)}^2$.
