---
title: "Transforming the Response"
institute: "Jam: **Stabilise** by Nilüfer Yanya"
---

```{r}
#| include: false
set.seed(2112)
```

::: {.content-visible unless-profile="book"}
## Preamble

### Announcements

- Check discussions!
:::

## Transformations

### Transforming the Predictors

Suppose we found that the following second order polynomial model was a "good" fit:
$$
y_i = \beta_0 + \beta_1x_{i1} + \beta_{11}x_{i1}^2 +\beta_2x_{i2} + \beta_{22}x_{i2}^2 + \beta_{12}x_{i1}x_{i2} + \epsilon_i
$$

\pause Now consider the model:
$$
\ln(y_i) = \beta_0 + \beta_1x_{i1} + \beta-2x_{i2} + \epsilon_i
$$

- 3 parameters instead of 6!
    - No interaction term!
- If we're okay with the log scale for $y$, easier to interpret.

### Transforming the predictors

:::: {.columns}
::: {.column width="50%"}
Original scale 
```{r}
#| echo: false
#| eval: true
#| fig-width: 5
#| fig-height: 5
library(plot3D)
library(ggplot2)
set.seed(2112)
n <- 200
x1 <- runif(n, 3, 10)
x2 <- runif(n, 3, 10)
y <- 20 + 5*x1 + 15*x1^2 + 5*x2 + 15*x2^2  + 2*x1*x2 + rnorm(n, 0, 2)

scatter3D(x1, x2, y, phi = 20, theta = 45, bty = "g",
    xlab = "x1", ylab = "x2", zlab = "y")
```

:::
::: {.column width="50%"}
Log scale
```{r}
#| echo: false
#| eval: true
#| fig-width: 5
#| fig-height: 5
scatter3D(x1, x2, log(y), phi = 20, theta = 45, bty = "g",
    xlab = "x1", ylab = "x2", zlab = "ln(y)")
```
:::
::::

The logarithm made it more linear, even if it's not quite right.

### Consequences of Logarithms

Consider the simple model $E(y_i) = \beta x^2$. Taking the logarithm of both sides:
$$
\ln(E(y_i)) = \ln(\beta) + 2\ln(x) = \beta_0 + \beta_1 \ln(x)
$$
and we have something that looks more like a linear model.

\pspace

- Note that, instead of $x^2$, $x^{2.1}$ would also work as a model.
    - The power of $x$ can be estimated.\lspace
- It's also possible that the log scale is the *correct* scale for $y$
    - $E(\ln(y_i)) = \beta_0 + \beta_1x$
    - In other words, don't get too bogged down by whether we take the ln of $x$.


### Logarithms and Errors

If we believe that the log scale is a better scale for $y$, we may postulate the model:
$$
\ln(y_i) = \beta_0 + \beta_1\ln x_{i1} + \beta_2\ln x_{i2} + \epsilon
$$
which implies that the orginal scale for $y$ has the form:
$$
y_i = e^{\beta_0}x_{i1}^{\beta_1}x_{i2}^{\beta_2}e^{\epsilon_i}
$$
*The errors are multiplicative!!!*

- Option 1: Accept this
    - Allows us to use least squares.\lspace
- Option 2: Use the model $y_i = e^{\beta_0}x_{i1}^{\beta_1}x_{i2}^{\beta_2} + \epsilon_i$
    - Might be better, but requires a bespoke estimation algorithm.

### General Practice

We often simply use the model:
$$
\ln \underline y = X\beta + \underline \epsilon
$$
and do *everything* on the log scale.

- Simpler, but still useful.\lspace
- Good predictions of $\ln y_i$ can be transformed to good predictions of $y_i$.\lspace

\pspace

In general: Decide on a functional relationship between $f(y)$ and $X$, then use additive errors on the scale of $f(y)$.

This has *consequences*:

## Residuals in Transformed Space

### Variance Stabilization

The two main purposes of transformations:

1. Fit non-linear functional forms.\lspace
2. Stabilize the variance!
    - Scale-Location plot in the R defaults.

\pspace

For example, the log function brings large values down a lot, small values down a little.

- The scale of large residuals is decreased more than the scale of small residuals.


### $f(\underline y) = X\beta + \underline\epsilon$

The estimated residuals are $\hat\epsilon_i = f(y_i) - \widehat{f(y_i)}$

- Note the awkwardly long hat! 
    - We're estimating the value of the function, not the value of $y_i$.
    - If $f(y_i) - \widehat{f(y_i)} = f(y_i - \widehat{y_i})$, then the original function must have been linear (and a transformation was useless).\lspace
- We're assuming $\epsilon_i\stackrel{iid}{\sim} N(0, \sigma^2)$, which is difficult to translate to $f^{-1}(X\beta + \underline\epsilon)$.
    - In the special case of $\ln$, $\exp{\epsilon_i} \sim \text{LogNormal}(0, \sigma^2)$.
    - No assumption of **independence** on the original scale!!!\lspace
- We assume that the residuals have the same variance on the *transformed* scale.
    - Likely not true for the original scale of $y$. 

### Some Good News

If $(a,b)$ is a $(1-\alpha)$ CI on the scale of $f(y)$, then $(f^{-1}(a), f^{-1}(b))$ is a valid CI on the scale of $y$.

\pspace

- It's not the only valid CI!
    - Note that it's not a symmetric CI!\lspace
- Works for $y$ as well as the $\beta$ parameters.
    - Transformation might induce dependence among the parameters.
    - A CI for $\beta_1$ is useless if there's high covariance with $\beta_2$. 

## Choosing Transformations

### Methods for Choosing Transformations

1. Theory.
    - If theory says that the log transform makes sense, use that.
        - Don't even consider the next steps. Just go with theory.
    - Example: Forest fire burn sizes are right skewed, the log-transform makes sense.
        - In my research, I used the lognorman lodel for the residuals to acheive the same effect.\lspace
2. Experimentation after looking at the Scale-Location plot.
    - If log or sqrt don't work, move on to step three.\lspace
3. The Box-Cox Transformation
    - Finds an appropriate transformation using maximum likelihood.

### Box-Cox

We use the transformation:
$$
V = \begin{cases}
\frac{Y^\lambda - 1}{\lambda \dot{Y}^{\lambda - 1}} & \text{if }\lambda \ne 0\\
\dot{Y}\ln(Y) & \text{if }\lambda = 0
\end{cases}
$$
where $\dot Y$ is the geometric mean of $y$.

\pspace

$\lambda$ is chosen through maximum likelihood

- Essentially, refit with each value of $\lambda$ and see which minimizes the residual variance.
    - Plot the likelihhods and choose the highest.

### Simpler Box-Cox

The textbook recommends the previous formula, however R uses:
$$
W = \begin{cases}
\frac{Y^\lambda - 1}{\lambda} & \text{if }\lambda \ne 0\\
\ln(Y) & \text{if }\lambda = 0
\end{cases}
$$

### Variance of $\lambda$

If we had a different data set, we'd get a different value of $\lambda$!

\pspace

R reports the the log-likelihood values, along with the top 5%.

- Anything in the top 5% is reasonable.
    - It's not an exact science.\lspace
- Usually, we check the best $\lambda$ values and round to something nice.
    - log, sqrt, squared, inverse, etc.

### Summary

- Choosing a transformation:
    1. Theory
    2. Exploration
    3. Round the value from Box-Cox.\lspace
- Working with a transformation:
    - Choose functional form, assume additive errors (usually, not always!)
    - Stay on the transformed scale
        - All assumptions about residuals apply to the transformed scale!

\pspace

To be useful, all transformations should consider the context of the problem!

::: {.content-visible unless-profile="book"}


## Participation 

### Q1

If the true relationship has the form $y = f(X) + \epsilon_i$, we can always find a transformation $f^{-1}$ to make it linear.

\pspace

1. True
2. False

### Q2

If we find that $\lambda = 2$ is the best transformation, then the following models are equivalent:
$$
\frac{Y^2 - 1}{2} = X\beta\quad\text{and}\quad Y^2 = 2X\beta + 1
$$

\pspace

1. True
2. False
3. Technically not, but good enough in practice.

### Q3

A transformation of the form $y = f(X) + \epsilon$ leads to multiplicative errors.

\pspace

1. True 
2. False

### Q4

Which of the following is *not* a good reason to investigate transformations?

\pspace

1. If the variance looks unstable.
2. If the theory says a transformation is necessary.
3. If a transformation might lead to a much simpler model.
4. If $y$ doesn't look normal.

### Q5

The default residual plots in R can help diagnose the need for a transformation.

1. True
2. False

:::


