---
title: "Hat Matrix and Residuals in R"
output: html_notebook
---


```{r}
#| include: false
set.seed(2112)
```


## The Hat Matrix

$$
H = X(X^TX)^{-1}X^T
$$

In R, we can calculate the diagonal of the hat matrix as follows:

```{r}
mylm <- lm(mpg ~ disp + wt, data = mtcars)
hatvalues(mylm) |> unname()
```

## Extracting the Diagonals from H

There isn't a built-in function for the full hat matrix (the diagonals are usually all you'll need). For demonstration, here are some demonstrations of the features of the hat matrix.

```{r}
X <- model.matrix(mylm)
H <- X %*% solve(t(X) %*% X) %*% t(X)
all.equal(diag(H), hatvalues(mylm))
```

## Features of H

In the code below, I use the `unname()` function because mtcars has rownames which make the output harder to see (this used to be the norm in R, but it's fallen out of fashion).

```{r}
colSums(H) |> unname()
rowSums(H) |> unname()
range(H) # -1 <= h_{ij} <= 1
range(diag(H)) # 0 <= h_{ii} <= 1
H %*% rep(1, ncol(H)) # H1 = 1
```

## Extracting the residuals

See `?influence.measures`.

```{r}
?rstandard
cooks.distance(mylm)
```

## The `broom` package

The `broom` package has a wonderful function called `augment()`. This function sets up our data so that it's super easy to see what we need.

```{r}
library(broom)
augment(mylm)
```

Notice that it includes:

-   `.fitted` = $X\underline{\hat\beta}$
-   `.resid` = $\hat{\underline\epsilon}$
-   `.hat` = $diag(H)$
-   `.sigma` = $s_{(i)}$
-   `.cooksd` = `D`
-   `.std.resid` = standardized or studentized residuals?

## Plotting the residuals

If you've ever accidentally typed `plot(mylm)`, you've seen some plots of the residuals.

```{r}
par(mfrow = c(2, 2))
plot(mylm) # ?plot.lm
```

You can access individual plots with the `which` argument.

```{r}
par(mfrow = c(2, 3))
plot(mylm, which = 1:6)
```

99% of the time, the default plots are the ones you'll want to look at. For teaching purposes, we'll look at a few extra.

## Demonstrations

We'll use the `Ozone` data from the `mlbench` package.

-   V4 is response (measurement of Ozone)
-   V5 is atmospheric pressure
-   V6 is wind speed
-   V7 is humidity
-   We'll ignore the rest.

```{r}
library(mlbench)
data(Ozone)
str(Ozone) # V4 is response (measurement of Ozone)
Ozone <- Ozone[complete.cases(Ozone), ]
```

A small amount of exploration first:

```{r}
library(ggplot2)
theme_set(theme_bw())
library(patchwork)
g5 <- ggplot(Ozone) +
    aes(x = V5, y = V4) +
    geom_point()
g6 <- ggplot(Ozone) +
    aes(x = V6, y = V4) +
    geom_jitter() # deal with overlapping points
g7 <- ggplot(Ozone) +
    aes(x = V7, y = V4) +
    geom_point()
g5 + g6 + g7
```

Now let's check some residuals!

-   Change `.resid` to `.std.resid`.
    -   Try `rstudent(olm)` as well.
-   Change `.hat` to `.cooksd`.

```{r}
library(dplyr)
olm <- lm(V4 ~ V5 + V6 + V7, data = Ozone)
augment(olm) %>%
    ggplot() +
        aes(x = .fitted, y = .std.resid, col = .cooksd) +
        scale_colour_viridis_c(option = 2, end = 0.7) +
        theme(legend.position = "bottom") +
        geom_point(size = 2) +
        geom_hline(yintercept = 0, colour = "grey")
```

Which ones have a large hat value?

The plots below are the same as the ones above, but coloured according to the hat values.

```{r}
g5 <- ggplot(Ozone) +
    aes(x = V5, y = V4, col = hatvalues(olm)) +
    geom_point(size = 2)
g6 <- ggplot(Ozone) +
    aes(x = V6, y = V4, col = hatvalues(olm)) +
    geom_jitter(size = 2) # deal with overlapping points
g7 <- ggplot(Ozone) +
    aes(x = V7, y = V4, col = hatvalues(olm)) +
    geom_point(size = 2)
(g5 + g6 + g7) +
    plot_layout(guides = "collect") &
    scale_colour_viridis_c(option = 2, end = 0.8) &
    theme(legend.position = "bottom")
```

## Adding an Outlier

Let's add an outlier to see what happens with these data.

```{r}
newzone <- Ozone[, c(4:7)]
newzone <- rbind(newzone,
    data.frame(V4 = 30, V5 = 5300, V6 = 5, V7 = 40))
newlm <- augment(lm(V4 ~ ., data = newzone))

g5 <- ggplot(newlm) +
    aes(x = V5, y = V4, col = .hat) +
    geom_point(size = 2)
g6 <- ggplot(newlm) +
    aes(x = V6, y = V4, col = .hat) +
    geom_jitter(size = 2) # deal with overlapping points
g7 <- ggplot(newlm) +
    aes(x = V7, y = V4, col = .hat) +
    geom_point(size = 2)
(g5 + g6 + g7) +
    plot_layout(guides = "collect") &
    scale_colour_viridis_c(option = 2, end = 0.8) &
    theme(legend.position = "bottom")
```

## Using R's Built-In Diagnostics

```{r}
par(mfrow = c(2, 2), mar = rep(2, 4))
plot(lm(V4 ~ ., data = newzone))
newzone[row.names(newzone) %in% c(1, 58, 243), ]
```

1.  Huge residual!
    -   This plot also just has a bad pattern
2.  Deviates from normality!
    -   Otherwise this looks pretty good.
3.  Large standardized residual
    -   Clear pattern without the outlier
4.  Cook's distance is massive compared to the others
    -   Potentially some large $D_i$'s
5.  In the corner
    -   Otherwise this looks okay-ish
6.  Last plot also shows it as something different (harder to interpret)

## What to do with a large residual?

-   Misrecorded: remove or fix, if possible
    -   Fires with negative lengths (MDY versus DMY)
    -   CO2 measured as -99 (code for NA in a system with no NA option)
    -   Heights measured in the wrong units
-   Real, but large residual: Consider whether it's actually part of the population of interest
    -   Studying heights and got a basketball player in your sample? That's a real data point and your model should allow for it!
    -   Studying fish and a shark was included? That's real, but maybe you should narrow your scope!
-   Many large outliers: you may need to try more predictors or a non-linear model.

Do *NOT* remove a point simply because it's an outlier!
